---
title: "Regional Differences in Health"
author: "Evangeline Warren"
date: "11/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (readstata13)
library (haven)
library (lattice)
library (stargazer)
```

## Overview
This project examines the relationship between skin tone and health and begins to interrogate how regional differences mediate this relationship. 

#### Data
The data comes from the GSS dataset and compiles all years with data for respondent's skin tone as assessed by the interviewer. It is available for download from the [GSS Data Explorer](https://gssdataexplorer.norc.org) site. 

I included the following variables:  
* R's Skin Tone (ratetone)  
* R's Self-Rated Health (health)  
* R's Days of activity limitation in the last month (hlthdays)  
* R's Days of poor mental health in the last month (mntlhlth)  
* R’s Health in general (health1)  
* Region (region)

## Packages
For this project, I used the following packages to ensrue I was able to read .dta files and so that I could make desired graphs. 

``` {r library}
#Setup
library (readstata13)
library (haven)
library (lattice)
library (stargazer)

```

## Cleaning
Data was cleaned in Stata (code in Stata folder) and then imported into R. I relabled the region data to have the correct labels and then relabelled health and health1 to have the same names for categories.  
```{r importing and cleaning}

#Data has been cleaned in stata and analytic sample has been restricted.
#Import cleaned data from .dta file

#Personal
Sample <-read_dta("~/Documents/GitHub/WarrenFinalGISD/Data and Analysis/ColorismCleaned.dta")
#View(Sample)

#Touch up the data to reflect qualitative Responses

Sample$region <- factor(Sample$region,
                    levels = c(1:9),
                    labels = c("New England"
                               , "Mid Atlantic"
                               , "E. Nor. Central"
                               , "W. Nor. Central"
                               , "South Atlantic"
                               , "E. Sou. Central"
                               , "W. Sou. Central"
                               , "Mountain"
                               , "Pacific"))


Sample$healthcl <- factor(Sample$healthcl,
                    levels = c(1:4),
                    labels = c("Excellent"
                               , "Good"
                               , "Fair"
                               , "Poor"))

Sample$health1cl <- factor(Sample$health1cl,
                          levels = c(1:4),
                          labels = c("Excellent"
                                     , "Good"
                                     , "Fair"
                                     , "Poor"))
```

## Initial Descriptive Plots

To understand the data, I made a histogram of the skin tone distribution. This revealed that the dataset is heavily skewed towards lighter skin tones

```{r histogram, echo=FALSE}
#Descriptive Plots
histogram(Sample$ratetonecl,
     xlab = "Assigned Skin Tone of R",
     main = "Distribution of Skin Tone in Sample",
     type = "count",
     breaks=c(1,2,3,4,5,6,7,8,9,10),
     col = c("purple"))
```

## Analytic Graphs

Because this project was partially comparitive (looking at national trends vs regional trends) I created two sets of graphs, ones that showed national trends and others that split based on region. 

#### Interval Level Variables for Health

First, I looked at the linear relationships between quantified values (number of poor mental health days, number of days of activity limitation). 

``` {r linears nat}
xyplot (Sample$mentalcl~Sample$ratetonecl,
        ylim=range(0:30),
        y.jitter=TRUE,
        x.jitter=TRUE,
        type = c("p", "r"),
        auto.key=TRUE,
        xlab = "Assigned Skin Tone of R",
        ylab = "Days of Poor Mental Health")

xyplot (Sample$healthdayscl~Sample$ratetonecl,
        ylim=range(0:30),
        y.jitter=TRUE,
        x.jitter=TRUE,
        type = c("p", "r"),
        auto.key=TRUE,
        xlab = "Assigned Skin Tone of R",
        ylab = "Days of Activity Limitation")
```

I then did the same, but split the trend lines based on region. 

```{r linear reg}
xyplot (Sample$mentalcl~Sample$ratetonecl,
        ylim=range(0:30),
        y.jitter=TRUE,
        x.jitter=TRUE,
        type = c("p", "r"),
        group = Sample$region, auto.key=TRUE,
        xlab = "Assigned Skin Tone of R",
        ylab = "Days of Poor Mental Health")

xyplot (Sample$healthdayscl~Sample$ratetonecl,
        ylim=range(0:30),
        y.jitter=TRUE,
        x.jitter=TRUE,
        type = c("p", "r"),
        group = Sample$region, auto.key=TRUE,
        xlab = "Assigned Skin Tone of R",
        ylab = "Days of Activity Limitation")
```

#### Ordinal Level Variables for Health
Finally, I followed the same approach for categorical values such as self rated health. I used histograms to show the distribution of self rated health overall and then by skin tone.  It is important to note that I had to change Rate Tone to a factor so that the regional histogram would accurately label the subdivision when I compared individual skin tone self rated health in the second graph. 
```{r self rated health}
Sample$ratetonecl<-as.factor(Sample$ratetonecl)
histogram(Sample$healthcl,
          xlab = "Self Rated Health",
          main = "Distribution of Self Rated Health, Overall",
          type = "percent",
          col = c("purple"))
histogram(~healthcl | ratetonecl,data=Sample,
          xlab = "Self Rated Health",
          main = "Distribution of Self Rated Health, by Skin Tone",
          type = "percent",
          col = c("purple"))
```